The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
worked through Section 2 (except 'pruning')
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2544 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8ccfb2cdfda6b953addda71640d46027293c7b0b

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various different tree 
algorithms have been developed, all following the same general structure starting at 
the root of the tree which is simply the first node containing all available data:
\begin{enumerate}
\item For the data set in the current node a model is fitted by optimizing
some objective function that reflects the goodness of fit.
\item Among all available covariates one is selected as the split variable,
choosing the split point such that goodness of fit is maximized in the
resulting subgroups.
\item Steps 1 and 2 are repeated within each subgroup until some stopping
criterion is attained.
\end{enumerate}

The term ``model'' is used here in a very broad sense and encompasses not
only least-squares or maximum-likelihood models but also simple constant
fits such as means or average proportions.
Therefore, depending on the type of model, the employed objective function can 
for example be the sum of deviations from a typical or average value, but it can 
also be constructed based on the model fitted in each node and depend not only on 
the response but also on possible covariates. For instance, different 
types of residuals or their signs can be applied \citep[see e.g.,][]{Loh:2002} as 
well as rank sums or logrank scores as in \cite{Hothorn+Hornik+VanDeWiel:2006}.

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, 
MOB, and GUIDE first select the split variable and then, in a separate step, the 
split point. For the first step of selecting a split variable they all apply statistical 
tests following the same general strategy: 
\begin{enumerate}
\item To capture how the objective function changes with the observations in
the current subgroup, a disaggregated, observation-wise goodness-of-fit
measure is obtained. More formally, this is an $n \times k$ matrix $G$
where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
the number of measures per observation.
\item The dependency or association of this goodness-of-fit matrix $G$ with
each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
using some suitable test statistic. The corresponding $p$-values allow for
a comparison of all $m$ covariates on a standardized or unified scale and in
that way for an unbiased split variable selection.
\item The covariate corresponding to the smallest $p$-value -- and thus the
highest influence on the goodness of fit of the model -- is selected for 
splitting the data into subgroups.
\end{enumerate}

As explained in more detail below, the aggregated overall objective
function in the linear regression model would be the residual sum of
squares and the corresponding observation-wise goodness-of-fit measure is
given by the residuals or scores (gradient contributions). Analogously,
the log-likelihood and corresponding score function could be used.

While this basic approach is the same for the GUIDE, the CTree and the MOB algorithm they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and $g(Z)$ for a so-called influence function $g$ and each possible split 
variable~$Z$.
The function $h$ has to depend on $Y$ in a permutation-symmetric way but can be of various forms
and may also be multidimensional. For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can return a location and scale parameter of $Y$, 
e.g. $h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fit to the values of the response variable~$Y$ and regressor 
variable(s)~$X$ the function $h$ can be a model-based transformation returning the 
corresponding model scores. This is done by setting $h(Y) = s(Y,X,\hat{\beta})$
where $\beta$ is the possibly multidimensional model parameter and $s$ is the score function
$$
s(Y, X, \beta) = \frac{\partial \psi(Y, X, \beta)}{\partial \beta}
$$
with the sum of $\psi$ over all observations being the aggregated overall objective 
function that is optimized to obtain the estimated model parameter $\hat{\beta}$. 
For this purpose many different M-type estimators can be applied including maximum 
likelihood or ordinary least squares such that the corresponding objective functions 
can also be employed for the tree algorithm. Evaluating the score function $s$ on a 
data set of $n$ observations 
%containing the 
%response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$ and 
%regressor(s) $\boldsymbol{x} = (x_1,\ldots,x_n)^\top$ and on
and a $k$-dimensional estimated parameter vector 
%$\hat{\boldsymbol{\beta}} = (\hat{\beta_1},\ldots,\hat{\beta_k})^\top$ 
yields a $n \times k$-matrix.
In that way a score value indicating how well the model fits the data is given separately
for each tuple of observation and estimated parameter element 
$(y_i, x_i, \hat{\beta_l})$ with $i \in \{1,\dots,n\}$
and $l \in \{1,\ldots,k\}$.

@@

To test for independence of $h(Y)$ and $g(Z)$ a linear test statistic is calculated.
Following \cite{Strasser+Weber:1999} the conditional expectation and covariance of the linear
test statistic given all permutations of the response variable can be calculated and used to 
standardize the test statistic. In order to map multivariate test statistics into the real 
line two options are given: either the maximum of the absolute values of the standardized 
statistic or a quadratic form. The resulting values can then be transformed to 
the scale of $p$-values allowing for an unbiased comparison in order to select a split variable.

If both variables $Y$ and $Z$ are numeric the applied independence test corresponds to a Pearson 
correlation test. For one of the variables being categorical and the other numeric a 1-way ANOVA
is employed while for two categorical variables a $\chi^2$-test is performed. Hence, the CTree 
algorithm provides a very general framework where the type of variables and the corresponding 
functions $g$ and $h$ determine the type of the applied statistical test.
(See \cite{Hothorn+Hornik+VanDeWiel:2006} for examples on how to use the different options of this toolbox.)

In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
observed values or optionally at an average value lying between two observations. To select the
best split point two-sample linear statistics of the same structure as in the independence test 
are evaluated for all possible subsets of the sample space of the selected split variable.

@@

While CTree offers to apply model-based goodness-of-fit measures as one possible choice, the 
MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was developed with the aim of embedding 
parametric models into the framework of regression trees and is hence purely intended for 
a model-based approach.

The testing strategy for selecting a split variable is based on a fitted model, i.e. on 
an estimated parameter (vector) $\hat{\beta}$ for each node of the tree, and on the 
application of the central limit theorem. 
The fitted model can be of a very basic and simple form such as an average value of the response 
vector but can also be a more complex parametric model.
In \cite{Zeileis+Hothorn+Hornik:2008} generalized linear models (GLMs) and survival regression models 
are used to exemplify the algorithm. Subsequently, various other types of models have been applied as well,
for example beta regression models \citep{Gruen+Kosmidis+Zeileis:2012}, 
psychometric models such as Rasch trees \citep{Strobl+Kopf+Zeileis:2015} and
mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.

Regardless of the applied model type the algorithm only requires the corresponding model 
score matrix in order to test for any instability in the model parameters over each 
particular ordering induced by each of the possible partitioning variables $Z$.
If a variable $Z$ has no specific influence on the parameters the ordered scores would 
always fluctuate randomly around their mean 0. Therefore, these deviations are investigated 
by measuring them using an empirical fluctuation process. Applying a scalar function which 
captures the fluctuation in this process yields the test statistic of this so called 
generalized M-fluctuation test \citep{Zeileis+Hornik:2007}. 
To obtain the distribution of the resulting test statistic asymptotic theory, in particular 
the central limit theorem is applied. In this parametric unconditional framework $p$-values 
can be calculated and the covariate with the smallest $p$-value is selected as split variable.

For a categorical split variable all possible splits are evaluated to find the best one while 
for numeric split variable a maximally selected split point search is employed. Hence, 
the number of possible breakpoints is not fixed unlike in the CTree algorithm.

\subsection{GUIDE}\label{sec:guide}
Similar to the MOB algorithm, parametric models are fit to the nodes of a tree in 
the GUIDE algorithm \citep{Loh:2002}. Again, simple models such as 
a constant fit with the parameter $\beta$ being the mean of the response, hence 
$\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$, 
%for a response vector $\boldsymbol{y} = (y_1,\ldots,y_n)^\top$, 
can be employed as well as more complex models.
For the statistical test applied to select a split variable some type of model-based
residual is required. In \cite{Loh:2002} linear models are used as an example where the
residuals are defined by
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In related literature other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002} 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015} among many others.
%Poisson regression trees \citep{Chaudhuri+Lo+Loh:1995}, %(adjusted Anscombe residual (Pierce and Schafer 1986))

Before applying a statistical test two additional steps are performed:
(1)~The residuals are binarized, i.e. depending on whether the sign of a residual is 
positive or not it is replaced by $1$ or $2$.  % FIX ME: +1 and -1 in the implementation? 
(2)~For each possible split variable the observed values are categorized into four groups 
separated by the corresponding 1st, 2nd and 3rd quartile unless the split variable is already
categorical.
Then, a $\chi^2$-test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
The possible number of breakpoints is determined by the categorization of the possible 
split variables and is in that way fixed.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$-test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

@@

\textit{Note:}
Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
statistic they apply and their approximation of the underlying null distribution of the corresponding 
test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
and GUIDE ($\chi^2$-tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
introduced employing prepruning while GUIDE originally applies a postpruning strategy.
However, these differences refer to the standard versions of the three algorithms and all of them
allow for certain modifications of this standard setup. For example, a different form of test statistic
can be selected in CTree considering a maximum value as in MOB.

@@

A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach one can consider residuals defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\psi$ is then
$$
\psi(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \psi(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
the score function 
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \psi}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\psi$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\psi$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.

>>>>

\subsection{Generic algorithm}
\label{sec:generic_alg}
The basic idea of building a regression tree model is to partition the data into smaller 
and more homogeneous subgroups based on a set of covariates. Various tree 
algorithms have been developed, following essentially the same general structure.
Starting at the root of the tree, pertaining to the full available data sample, the
algorithms proceed in the following steps:
%
\begin{enumerate}
  \item For the current sample a (possibly simple) model is fitted by optimizing
        some objective function (or loss function) that reflects the goodness of fit.
  \item Among all available covariates one is selected as the split variable,
        choosing the split point such that goodness of fit is maximized in the
        resulting subgroups.
  \item Steps 1 and 2 are repeated within each subgroup until some stopping
        criterion is attained.
\end{enumerate}
%
The term ``model'' is used here in a very broad sense and encompasses not
only least-squares or maximum-likelihood models but also simple constant
fits such as means or average proportions.
Therefore, depending on the type of model, the employed objective function can 
for example be the sum of deviations from a typical/average value, but it can 
also be based on a model fro a response along with potential covariates.
For instance, different types of residuals (or the signs thereof) can be employd
\citep[see e.g.,][]{Loh:2002} as well as rank sums or logrank scores
\citep[as in][]{Hothorn+Hornik+VanDeWiel:2006}.

As explained in Section~\ref{sec:introduction} the considered tree algorithms CTree, 
MOB, and GUIDE first select the split variable and then, in a separate step, the 
split point. For the first step of selecting a split variable they all apply statistical 
tests following the same basic strategy: 
%
\begin{enumerate}
  \item To capture how the objective function changes with the observations in
        the current subgroup, a disaggregated, observation-wise goodness-of-fit
        measure is obtained. More formally, this is an $n \times k$ matrix
        where $n \in \mathbb{N}$ is the number of observations and $k \in \mathbb{N}$ 
        the number of goodness-of-fit measures per observation.
  \item The dependency or association of this goodness-of-fit matrix with
        each possible split variable $Z_j$, $j \in \{1, \dots, m\}$, is assessed
        using some suitable test statistic. The corresponding $p$-values allow for
        a comparison of all $m$ covariates on a standardized or unified scale and in
        that way for an unbiased split variable selection.
  \item The covariate corresponding to the smallest $p$-value -- and thus the
        highest influence on the goodness of fit of the model -- is selected for 
        splitting the data into subgroups.
\end{enumerate}
%
As an example -- and explained in more detail below -- consider a linear regression tree.
Thus, a linear regression model is fitted in each subgroup, minimizing the
residual sum of squares as the aggregated goodness-of-fit measure. The corresponding
observation-wise goodness-of-fit measure can be given by the residuals or the scores
(gradient contributions). Analogously, the log-likelihood and corresponding score function could be used.

While this basic approach is the same for the GUIDE, CTree, and MOB algorithms, they differ 
in their strategies on how to calculate test statistics. In order to point out these specific 
characteristics in Section~\ref{sec:unify} the strategies of the three tree algorithms are first 
explained in more detail in Sections~\ref{sec:ctree}, \ref{sec:mob}, and \ref{sec:guide}.

@@

The CTree algorithm \citep{Hothorn+Hornik+Zeileis:2006} is based on the idea of providing 
non-parametric regression tree models in a conditional inference framework by applying 
permutation tests. 
To select a split variable it is tested whether there is any association between the transformed 
response $h(Y)$ and each possible transformed split variable $g(Z_j)$ ($j = 1, \dots, m$).
The only requirement for the function $h$ is to depend on $Y$ in a permutation-symmetric way but
this encompasses ranks, scores, indicator functions, etc.\ and can also be multidimensional.
For a numeric response the identity function $h(Y) = Y$ is a common 
choice while a categorical response can be mapped to a unity vector by an indicator function 
$h(Y) = (0,\ldots,1,\ldots,0)^{\top}$.
Alternatively, the function $h$ can capture location and scale of $Y$ via
$h(Y) = (Y, (Y-\bar{Y})^2)^{\top}$. 
If a parametric model is fitted to the response $Y$ with regressor variable(s)~$X$,
then a model-based transformation $h(Y) = s(Y,X,\hat{\beta})$ can be used, e.g.,
the residuals in a linear model with regression coefficients $\beta$. Moreover,
$s$ can be the score function pertaining to the objective function (or loss function)
$\ell$:
$$
s(Y, X, \beta) = \frac{\partial \ell(Y, X, \beta)}{\partial \beta}
$$
The estimate of the model parameters $\hat{\beta}$ is obtained by optimizing
the sum of the objective function $\ell$, aggregated over all observations.
This framework includes many different M-type estimators as special cases, including maximum 
likelihood and ordinary least squares estimation. For a $k$-dimensional parameter
$\beta$ the score function evaluated for the $i$-th observation $s(y_i, x_i, \beta)$
is also a $k$dimensional vector, i.e., the gradient contribution of the $i$-th observation.
Thus, the $n \times k$-matrix consisting of these scores or gradient contributions for all 
$i = 1, \dots, n$ observations is a natural candidate for the observation-wise goodness-of-fit
measure as described in the previous Section~\ref{sec:generic_alg}.

@@

To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
covariance given all permutations of the response variable can be calculated and used to 
standardize the test statistic. This standardized statistic has an asymptotic normal distribution
which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
multivariate. The actual test is carried out be mapping this standardized statistic to
the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
being computed from the analogous transformation of the normal distribution
\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.

If both variables $Y$ and $Z$ are numeric the default independence test corresponds to a Pearson 
correlation test. For one numeric and one categorical variable essentially a 1-way ANOVA
(analysis of variance) is employed while for two categorical variables a $\chi^2$~test is performed.
Thus, in the general CTree framework many types of tests can be specified by selecting suitable
transformations $g$ and $h$. While originally
conceived for nonparametric models, it is easy to adapt CTree to model-based testing and 
recursive partitioning by choosing a model-based $h$ transformation as argued above
\citep[see the concrete examples in][]{Zeileis+Hothorn:2013,Seibold+Zeileis+Hothorn:2016}.

% In the CTree algorithm the number of possible breakpoints is fixed as splits can only be made at
% observed values or optionally at an average value lying between two observations. To select the
% best split point two-sample linear statistics of the same structure as in the independence test 
% are evaluated for all possible subsets of the sample space of the selected split variable.

@@

In contrast to CTree the MOB algorithm \citep{Zeileis+Hothorn+Hornik:2008} was explicitly
designed for a model-based goodness-of-fit measure in order to embed parametric models
into a regression tree framework. Thus, MOB is based on an objective/loss function $\ell$
and corresponding score function $s$. The original paper considered generalized linear models
(GLMs) and survival regression models but subsequently various other models have been applied as well,
including beta regression \citep{Gruen+Kosmidis+Zeileis:2012}, psychometric item response theory
models \citep{Strobl+Kopf+Zeileis:2015}, or mixed effects models \citep{Fokkema+Smits+Zeileis:2018}.
But just like CTree can be applied to parametric models, MOB conversely also encompasses
simple regression and classification trees, e.g., by choosing an intercept-only model.

For selecting a split variable MOB employs a score-based test that relies on the central limit
theorem for the parameter estimate $\hat{\beta}$. The test assesses whether the scores
-- when ordered by the potential split variable $Z$ -- fluctuate randomly around their
zero mean or differ systematically in certain subgroups. The latter would indicate a
parameter instability that could be captured by fitting separate models (optimizing $\ell$)
in the resulting subgroups. In case of a numeric split variable $Z$ both the score-based statistic
and the partitioned objective function $\ell$ are maximized over all possible splits in $Z$
(subject to certain minimal subgroup size constraints). Unlike the CTree framework, MOB
relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007}.

\subsection{GUIDE}\label{sec:guide}

Building on earlier work for the QUEST algorithm \citep{Loh+Shih:1997}, \citep{Loh:2002}
proposed the GUIDE algorithm blending trees with parametric regression models and
encompassing simpler classification and regression trees as special cases. Thus,
linear regression models could be fitted in the nodes of tree as well as
constant fits such as simple mean response $\hat{\beta} = \bar{y} = \frac{1}{n}\sum_{i=1,\ldots,n}^n y_i$,.
The tests for selecting the split variable are then based on the corresponding residuals,
e.g., for a simple linear regression \citep[as used in][]{Loh:2002}:
$$
r(Y,X,\hat{\beta}) = Y - \hat{\beta_0} - \hat{\beta_1} \cdot X.
$$
In subsequent work other models together with an appropriate choice of residuals
have been applied such as in
regression trees for longitudinal and multiresponse data \citep{Loh+Zheng:2013}, 
%(http://www.stat.rice.edu/~jrojo/4th-Lehmann/slides/Loh.pdf)
quantile regression models \citep{Chaudhuri+Loh:2002}, 
and proportional hazards modeling via Poisson regression \citep{Loh+He+Man:2015}, among others.

To construct a statistical test two additional transformation are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~All possible split variable $Z$ are categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test.

@@

\item Categorization of covariates:\\
Similarly, the covariates can also be categorized. For this purpose, the corresponding 1st, 2nd 
and 3rd quartile are considered to bin each possible split variable into four categories. In that way,
depending on whether the scores are binarized or not, a $\chi^2$~test or 1-way ANOVA can be applied
where the quartiles are the only possible split points while otherwise maximally selected splits 
over $Z$ are considered.

@@

\textit{Note:}
Apart from these three building blocks, the algorithms CTree, MOB, and GUIDE differ in the type of test 
statistic they apply and their approximation of the underlying null distribution of the corresponding 
test statistic. While the permutation tests in CTree work in a conditional framework both MOB (M-fluctuation tests) 
and GUIDE ($\chi^2$~tests) apply an unconditional approach. Furthermore, the algorithms CTree and MOB were 
introduced employing prepruning while GUIDE originally applies a postpruning strategy.
However, these differences refer to the standard versions of the three algorithms and all of them
allow for certain modifications of this standard setup. For example, a different form of test statistic
can be selected in CTree considering a maximum value as in MOB.

@@

A very basic form of goodness-of-fit measure can be obtained for each observation by simply 
calculating deviations of the response values from the mean value within the node, for example 
by $(\bar{y} - y_i)$.
However, this includes only the response variable $Y$ but not the fitted linear model.
For a model-based approach one can consider residuals defined by
$$
r(y_i,x_i,\beta_0,\beta_1) = y_i - \beta_0 - \beta_1 \cdot x_i.
$$
The appropriate choice of the function $\ell$ is then
$$
\ell(y_i,x_i,\beta_0,\beta_1) = r(y_i,x_i,\beta_0,\beta_1)^2
$$
such that $\sum_{i=1}^n \ell(y_i,x_i,\beta_0,\beta_1)$ is the objective function
which is minimized to obtain the estimated coefficient vector
$\boldsymbol{\hat{\beta}} = (\hat{\beta_0}, \hat{\beta_1})^\top$.
The corresponding disaggregated, observation-wise goodness-of-fit measure is defined by
the score function 
$$
s(y_i, x_i, \beta_0, \beta_1) = 
\frac{\partial \ell}{\partial \boldsymbol{\beta}}(y_i, x_i, \beta_0, \beta_1) = 
-2 \cdot r(y_i,x_i, \beta_0, \beta_1) \cdot (1,x_i)^\top
$$
where the factor $-2$ is often omitted such that the first column of the $n \times 2$-dimensional
score matrix corresponds to the residuals. 
Therefore, including only residuals (i.e. only the first column of the score matrix)
corresponds to including the derivatives of $\ell$ only regarding the intercept 
parameter~$\beta_0$.
On the contrary, by using full model scores (i.e. the whole $n \times 2$ matrix) 
the derivatives of $\ell$ regarding the intercept $\beta_0$ (first column) and the slope 
parameter $\beta_1$ (second column) are considered.
